\mytitle

### 2023-01-31 paper-seb-main.tex
Fixing git messages

\PassOptionsToPackage{hyphens}{url}
\input plos-header.tex

\renewcommand\thesubsection{(\thesection.\alph{subsection})}
\usepackage{hyperref}

\def\httpURL#1{\href{http://#1}{\textcolor{blue}{#1}}}
\def\LONGhttpURL#1#2{\href{http://#1}{\textcolor{blue}{#2}}}

\makeatletter
    \def\do@url@hyp{\do/}
    \usepackage{moreverb,url}
    %    \renewcommand{\thefootnote}{\arabic{footnote}}
\makeatother

\input paper-seb-macros.tex

\input generated-page-lengths.tex

@@

\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\mytitle} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Harold Thimbleby
\\
\bigskip
See Change Fellow in Digital Health
\\
\bigskip

% Current address notes

% Use the asterisk to denote corresponding authorship and provide email address in note below.
\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
&\texttt{harold@thimbleby.net}\\
&ORCID 0000-0003-2222-4243
\end{tabular}

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
\noindent
\emph{Background:} Computer code underpins modern science. While scientific models are routinely peer reviewed, the algorithms and the quality of code implementing them often avoid scrutiny; therefore the details of many scientific conclusions cannot be rigorously justified.

\noindent\emph{Problem:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit. Additionally, both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is therefore fallible, and if the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

\noindent\emph{Methodology:} This paper samples and reviews software engineering practice in peer reviewed published in leading scientific journals.

\noindent\emph{Results:} On the basis of the sample evidence, scientists rarely assure the quality of any code they rely on, rarely make complete code available, and rarely provide adequate documentation to use code reliably.

\noindent\emph{Solutions:} Code can be improved using software engineering. This paper argues for specific solutions:

\newdimen \mywidth \mywidth=\textwidth
\advance \mywidth by -1em
\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 1.\hfill}Professional software engineers can help and should be involved, particularly in critical research;

\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 2.\hfill}``Software Engineering Boards'' (analogous to Ethics or Institutional Review Boards) should be instigated and used;

\parshape 2 0em \textwidth 1.5em \mywidth 
\noindent 
\hbox to 1.5em{ 3.\hfill}Code, when used, should be considered an intrinsic part of any publication, and therefore should be formally reviewed by competent software engineers.

\noindent
The \supplement\ provides a summary of professional software engineering best practice relevant to scientific research and publication.

%\end{abstract}

\vfill\begin{quote}
\hfill``Criticism is the mother of methodology.''

\hfill Robert P Abelson, \emph{Statistics as Principled Argument\/} \cite{abelson}\end{quote}
\vfill
\paragraph*{Keywords:} 
Computational science; Software Engineering; Reproducibility; Scrutiny

\newpage\section*{Author summary}

%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%In addition to over 340 peer reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.
\linenumbers

%\maketitle

\section{Introduction}\label{problems}
Computational methods are comparatively new, powerful, and very flexible tools for scientists. Computational methods rely on programming, but unfortunately poor programming is very easy use to apparently get results with; such results are likely to be misleading. The role of computational methods can be compared to statistics, another useful scientific tool. Poor statistics is much easier to do than good statistics, and there are many examples of science being let down by na\"\i vely planned and poorly implemented statistics. Often scientists do not realize the limitations of their own statistical skills, so careful scientists generally work closely with professional statisticians.

{I}{n good science}, all statistics, methods and results have to be reported very carefully and in precise detail. For example, a statistical claim might be summarized as follows:

\begin{quote}\raggedright
\setbox0=\hbox{``}
\sf\hskip -\wd0\box0 %Random intercept linear mixed model suggesting significant time by intervention-arm interaction effect. 
\ldots\ Bonferroni adjusted estimated mean difference between intervention-arms at 8-weeks 2.52 ~($\sf 95\% \mbox{\sf ~CI~} 0.78, 4.27, p = 0.0009$). Between group effect size $\sf d = 0.55 ~(95\% \mbox{\sf ~CI~} 0.32, 0.77)$.'' \cite{example-stats}\end{quote}

This typical wording briefly summarizes confidence intervals, $p$ levels, and so on, to present statistical results so the paper's claims can be seen to be complete, easy to interpret, and easy to scrutinize. It is a \emph{lingua franca}. It may look technical, but it is written in the standard and now widely accepted form for summarizing statistics.

Scientists write like this --- and conferences and journals require it --- because statistical claims need to be properly accountable and documented in a clear way. Speigelhalter \cite{Speigelhalter} says statistical information needs to be accessible, intelligible, assessable, and usable; he also suggests probing questions to help assess statistical quality (see \supplement\ section~\ref{supplementary-Speigelhalter-section}). Results should not be uncritically accepted just because they are claimed.

The skill and effort required to do statistics so it can be communicated clearly and correctly, as above, is not to be taken for granted. Scientists work closely with competent, often specialist, statisticians who engage with the research from experiment design through to analysis and publication. Further, it is assumed that statistics will be peer reviewed, and that review will often lead to improvement.

>>>>

\documentclass{comjnl}
\usepackage{ragged2e}
\usepackage{amsmath}
\usepackage{amssymb}

%----
  \setlength\textheight{220mm}
%----

\input paper-seb-macros.tex
\input generated/info-for-main.tex
\input generated/page-lengths.tex

\long\def\ignore#1{}

\title{\mytitle} 
\author{Harold Thimbleby}
\shortauthors{Harold Thimbleby}
\affiliation{See Change Fellow in Digital Health, Swansea University}
\email{harold@thimbleby.net}

%\subject{Computational science (bioinformatics,~medical~computing,~etc)}

\keywords{Computational Science; Software Engineering; Reproducibility; Scientific scrutiny; Reproducible Analytic Pipeline (RAP \& \RAPstar\hskip -1pt)}

\received{30 April 2022}
%\revised{--}

%\begin{tabular}{@{}ll}Current Address: &62 Cyncoed Road, Cardiff, CF23 5SH, Wales\\
%&\texttt{harold@thimbleby.net}\\
%&ORCID 0000-0003-2222-4243
%\end{tabular}

@@

\begin{abstract}
\justifying
%\color{red}
%This is a placeholder for the revisions that are almost complete. I have uploaded it now in case the computer system would lose all record of the submission --- I have emailed to ask for a short extension, but with the offices shut, there is no chance of any negotiation. Thanks for your understanding.
%\color{black}\vskip .2cm

\noindent
Scientists rarely assure the structure and quality of the code they rely on, and they rarely make it accessible for scrutiny. Even if available, scientists rarely provide adequate documentation to understand or use their code reliably. When code used in science avoids adequate scrutiny published science will be unreliable, because results --- data, graphs, images, inferences, etc --- generated by the code will themselves be unreliable.

\hskip 1em This paper justifies and proposes ways to improve research and science relying on code:

%\emph{Problems:} Assumptions in scientific code are hard to scrutinize as they are rarely made explicit, even when code is made available. Both algorithms and code have bugs, unknown and accidental assumptions that have unwanted effects. Code is fallible, so any interpretation that relies on code is also fallible. When the code is not clearly structured and published with adequate documentation, the code cannot be usefully scrutinized. In turn, scientific claims cannot be properly scrutinized.

%\emph{Methodology:} From the perspective of Software Engineering, this paper critiques the quality (and accessibility) of coding in computational science, particularly as relied on in leading COVID-19 pandemic research driving international public health interventions. 
%The paper reports a pilot survey of peer-reviewed computational modeling papers ($N=\the\dataN$) published in leading scientific journals.

%Code can be improved using Software Engineering. This paper argues for specific solutions:

%\emph{Results:} %Journals do not always have a code policy, and many have a relaxed approach to code disclosure.

%\noindent\emph{Solutions:}

\raggedright
\newdimen \mywidth \mywidth=\textwidth
\advance \mywidth by -3em

\def\indented#1{\vskip 1ex
\parshape 2 0em \textwidth 2em \mywidth 
\noindent 
\hbox to 2em{ #1.\hfill}\ignorespaces}

\indented 1
Professional Software Engineers can help, particularly in critical fields such as public health, climate change, and energy.

\indented 3
``Software Engineering Boards,'' analogous to Ethics or Institutional Review Boards, should be instigated and used.

\indented 4
The Reproducible Analytic Pipeline (RAP) methodology can be generalized to cover code and Software Engineering methodologies, in a generalization this paper introduces called \RAPstarp. \RAPstar\ (or comparable interventions) could be supported and encouraged in journal, conference, and funding body policies.

\vskip 1ex 
\justifying
\noindent
The key point is that as code is now an inextricable part of science it should be supported by competent Software Engineering input, analogously to statistics being properly supported by critical and competent statistics input. The \supplement\ provides a summary of Software Engineering best practice relevant to scientific research, including further suggestions for \RAPstar\ processes.

%\begin{quote}
\def\seprule{\vskip 2ex\hrule\vskip 2ex}
\seprule
\raggedright
\def\zq{\setbox0=\hbox{``}\hskip -0\wd0\copy0}
\zq Science is what we understand well enough to explain to a computer.''\\
\hfill {\rm Donald E. Knuth in $A=B$ \cite{a=b}}\\

\zq I have to write to discover what I am doing.''\\
\hfill {\rm Flannery O'Connor, quoted in \emph{Write for your life\/} \cite{write-for-life}}

\zq Criticism is the mother of methodology.''\\
\hfill {\rm Robert P. Abelson in \emph{Statistics as Principled Argument\/} \cite{abelson}}

\zq \hangindent \wd0 From its earliest times, science has operated by being open and transparent about methods and evidence, regardless of which technology has been in vogue.''\\
\hfill {\rm Editorial in \emph{Nature\/} \cite{nature-chatgpt}}

\seprule
%\end{quote}
\end{abstract}
\maketitle

%\section*{Author summary}
%
%%\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{HT}}]
%Harold Thimbleby PhD, FRCP (Edinburgh), Hon.\ FRSA, Hon.\ FRCP is See Change Fellow in Digital Health at Swansea University, Wales. His research focuses on human error and computer system design, particularly for healthcare. 
%%In addition to over 340 peer-reviewed and 188 invited publications, Harold has written several books, including \emph{Press On\/} (MIT Press, 2007), which was winner of the American Association of Publishers best book in computer science award.
%Harold won the British Computer Society Wilkes Medal. He is emeritus Gresham Professor of Geometry (a chair founded in 1597), and has been a Royal Society-Leverhulme Trust Senior Research Fellow and a Royal Society-Wolfson Research Merit Award holder. His latest book is  \emph{Fix IT: How to see and solve the problems of digital healthcare\/} \cite{fixit}.  See his web site, \texttt{www.harold.thimbleby.net}, for more details.

%\noindent\emph{This paper evidences widespread problems arising from scientific papers relying on inadequately defined computer and inaccessible code, which undermines peer-review and replication, and undermines scientific quality generally. The risks vary, depending on how critical a paper's claims may be: ranging from superficial for opinion papers, to serious for papers in domains such as medicine, public health, or policy. This paper contrasts the situation with the mature approach to using statistics, and goes on to make proposals to help address the problem. The proposals are feasible, and presented in the hope of motivating their (or better solutions) being implemented, tested, and refined.}

\section{Introduction}
Unreliable, often unstated and unexplored, code and computational dependencies (including using AI or ML) in science are widespread. Furthermore, code is rarely published or made accessible in a useful form, nor is it professionally scrutinized. Too often, therefore, code and results claimed do not contribute to reliable science, and do not support replication or reproduction on which future science can be firmly based.

\begin{table}
\vskip 1.2ex % align top of columns!
\tableofcontents
\vskip 3mm \hrule
\end{table}

This paper makes an explicit analogy between the use of statistics, which has clear standards for reliable use and presentation, and the use of code. Like statistics, code is often relied on to support key results in scientific publications, yet code is generally informal, inaccessible and incorrect. Just as sound experimental methods and sound statistics generally rely on professional specialist input, it is argued here that good use of code must rely on professional Software Engineering input \cite{sommerville,knight}, and more strategically on Computational Thinking \cite{computational-thinking-wing,computational-thinking} --- an accessible form of Software Engineering that consciously applies the ideas universally, far more widely than just to software and coding.

\begin{table*}
\begin{center}\sf\normalsize
\begin{tabular}{|rl|} \hline
\the\numberOfJournals&Journals\\
\the\dataN&Papers:\\
\tabularJournalBreakdown
\the\countAuthors&Published authors\\
\the\totalPages&Published journal pages\\
July 2020&Sample month\\ \hline
\end{tabular}
\end{center}
\caption{Overview of peer-reviewed paper sample, broken down further in Table \ref{table-summary}. Survey methodology and data is provided in the \supplement. (The survey does not include the motivating papers \cite{ICmodel,avianFluModel,originalICmodel}, none of which provide code or code summaries; see section \ref{section-pandemic-modeling}.)}
\label{table-overview}
\end{table*}

\begin{table*}
\begin{center}\sf\normalsize
\input generated/summary-table.tex
\end{center}
\caption{Breakdown of pilot survey of peer-reviewed science papers relying on code. }
\label{table-summary}
\end{table*}

\begin{table*}
\newdimen\quarterwidth
\newdimen\separation
\separation=3mm
\quarterwidth=\textwidth
\advance\quarterwidth by -4\separation
\divide\quarterwidth by 4
\noindent{\sf\begin{tabular}{@{}c|c|c|c@{}}
\bfseries Level 0&
\bfseries Level 1&
\bfseries Level 2&
\bfseries Level 3 \\ \hline
\multicolumn{1}{@{}p{\quarterwidth}}{Journal encourages code sharing -- or says nothing.}&
\multicolumn{1}{|p{\quarterwidth}}{Article states whether code is available and, if so, where to access them.}&
\multicolumn{1}{|p{\quarterwidth}}{Code must be posted to a trusted repository. Exceptions must be identified at article submission.}&
\multicolumn{1}{|p{\quarterwidth}@{}}{Code must be posted to a trusted repository, and reported analyses will be reproduced independently before publication.}
\end{tabular}}
\caption{The TOP committee's recommended levels for journal article code transparency. Level 0 is provided for a comparison that does not meet any TOP requirements. Concerns about the interpretation of ``reproduced independently,'' as required at level 4, are raised in section \ref{reproducibility-concerns}.}
\label{TOPtable}
\end{table*}

This paper was initially motivated by concerns about the poor quality of code used in high-profile epidemiology research, because of its significance for informing and driving public health responses to the Covid pandemic. A pilot survey implies that such problems are ubiquitous and by no means limited to epidemiology: see the tables \ref{table-overview} \& \ref{table-summary} for a summary of the sample, and see the \supplement\ for further details of the survey. In the survey, \emph{no\/} papers claimed or provided evidence that their code was adequately tested or rigorously developed; \emph{none\/} used methodologies like RAP or \RAPstar\ (described below). Only one paper mentioned any Software Engineering methods, albeit without technical details.

In the survey sample 81\% of papers were published in journals that have code policies (which themselves are weak), but 42\% of surveyed papers published in those journals breached their own policies. One paper declared it had accessible code, but the relevant repository was and still remains empty. The findings are comparable to problems increasingly recognized for data and data access (reviewed in section \ref{related-work}): code problems form part of the \emph{reproducibility crisis\/} \citeeg{reproducibility-crisis,rescience}.

This paper therefore argues that code should be developed and discussed in a professional, rigorous, and supportive environment that facilitates quality science with clear presentation and appropriately rigorous scrutiny of code. Its main contribution is to suggest straightforward ways to enable this. The proposals may not be ``the'' right or best ways, but it is hoped the case studies and arguments presented here persuade readers that the proposals are at least a productive way to start pointing in the right direction, and inform raising the profile and constructively debating the issues more widely.

An extensive online \supplement\ appendix to this paper provides additional resources, including brief details of many Software Engineering practices relevant to supporting quality science. The supplement will be of particular interest to research software engineers supporting non-software-specialist scientists.

\section{Background}
The discoveries and inventions of scientific technologies and instruments like microscopes, telescopes, and X-rays, drive and expand the sciences. There are fascinating periods as new ideas and science unify; for example, thermometers could not measure temperature in any meaningful way until the science was mature. For over a century, there was no agreement on definitions of temperature, how to calibrate thermometers, or what units they measured in.

Paradoxically the science could not mature until there was consensus in scientific methodologies for thermometry, and having a consensus in turn depended on reproducible thermometer measurements and a thorough understanding of the science, including all the confounding factors that were being misunderstood \cite{temperature}.\footnote{One example: if the volume of mercury is chosen to measure temperature, a confounding factor is that the volume of the container measuring the mercury volume also increases with temperature (but at a different rate), so the volume measurement is inaccurate.} 
Science matured from no quantitative interest in temperature, through a complex process of hand-in-hand theoretical-and-technical maturation, until today, when we have robust off-the-shelf instruments that measure temperature in reliable, repeatable, internationally standardized units, following international quality standards.

Computers are a unique, new technology, far more flexible and challenging than thermometers. Understanding computers and integrating them into science is likely to be harder than the development of modern thermometry. Computation not only expands science's paradigms and supports new discoveries (particularly with AI), but it also \emph{does\/} new science --- almost all modern laboratory instrumentation, including thermometers, is heavily computerized. Pushing the boundaries of science, then, involves pushing the boundaries of computer science. The synergy runs deep: for instance, while particle physics relies on powerful supercomputers, quantum physics itself is developing more powerful quantum computing.

``Computational science'' has come to mean a particular style of science based on developing and using explicit computational models, but, really, \emph{all\/} of science is now computational in this sense. Computational science is not just restricted to specialized fields like computational chemistry, genomics, big data~\ldots\ In all fields of science, computation is used at every step, from calculations of course, through note taking, sound and image processing, literature searches, analysis and statistics, correspondence with co-authors and editors, through to typesetting,  distributing and archiving the final publications. All areas of science are being profoundly computerized. Furthermore, developments in computer science themselves drive science such that earlier science is even becoming obsolete as the computer technology moves on \cite{form}.

\subsection{Related code quality concerns}\label{related-work}
Publishing high quality computer code has been strongly advocated since the earliest times, such as the \emph{Communications of the ACM\/} in its first issue in its first volume published in 1958, where it outlined its algorithm publication policy. The new policy was illustrated with a square root algorithm \cite{acm-algorithms}. However, publishing code in the computer science literature is distinct from publishing high quality general science that depends on code, which is the concern of the present paper. Of course, as a special case, Computer Science too can also benefit from improving ways to reliably use code.

In almost all published science, the code it relies on is taken for granted, just as in routine chemistry the quality of the glassware is not at issue. While chemists are trained in reliable methodologies, so taking quality glassware for granted is reasonable. Code is a newer innovation. Reliable code is a current research programme in its own right, and has resulted in calls for a Grand Challenge research effort \cite{hoare-correctness}. Inevitably, because of these reasons taken collectively, much published science depends on unreliable code that is not explicitly discussed, was not peer-reviewed, and is not open to scrutiny, reproduction, or reuse by the scientific community.

A study of 863,878 Python-coded Jupyter notebooks \cite{jupyter-study} found a 76\% failure rate for code to complete execution successfully. Trisovic \emph{et al.}\ \cite{trisovic} performed a  study of 9,000 research codes written in the language R on Dataverse, an open-source repository maintained by Harvard University's Institute for Quantitative Social Sciences. They found a comparable result that 74\% of the code files analyzed failed. These results are consistent with this paper's findings, as summarized in tables \ref{table-overview} \& \ref{table-summary}.

The authors of \cite{jupyter-study} make technical recommendations to improve reproducibility, such as ``Abstract code into functions, classes, and modules and test them.'' More generally, the authors of \cite{trisovic} recommend establishing Working Groups to support reproducible research --- and in fact Dataverse has already done so. These ideas may be compared to the present paper's proposal of Software Engineering Boards (SEBs), as discussed in detail in section \ref{summary}, supported by more specific suggestions in the \supplement.

These concerns about code are part of the reproducibility crisis for science generally \cite{machine-learning-reproducibility,no-raw-data,data-access,reproducible-manifesto,reproducibility-crisis,reproducible-manifesto}. Concern has led to a new journal \emph{ReScience~C\/} to explore and encourage the explicit replication of previously published research \cite{rescience}; furthermore, \emph{ReScience~C\/} promotes new and open-source implementations in order to ensure that the original research is reproducible. However, most of its published concern focuses on data rather than code.

To start to address code quality issues, the Transparency and Openness Promotion Committee first met in 2014, and has since been promoting Transparency and Openness Promotion, TOP, starting with journal publication policies \cite{TOP}. TOP covers citation standards, replication standards, and code standards amongst others. TOP recommends levels of compliance to their recommendations, where level 0 does not meet the standard, and levels 1 to 3 are increasingly stringent. The TOP levels for code are shown in table \ref{TOPtable}. The TOP standards  continue to develop, and are now maintained on a wiki at \url{osf.io/9f6gx/wiki/Guidelines} \cite{TOP-wiki}. Note that TOP aligns with but is a stricter approach than the Findable, Accessible, Interoperable and Reusable (FAIR) initiative, which is critiqued in section \ref{critique-fair}.

As the present paper argues, developing quality code is widely under-appreciated, which leads to a vicious cycle of lack of acknowledgement, invisibility, and being unable to recruit adequately competent coders (section \ref{section-vicious-cycles}). The term \emph{research software engineer\/} was coined in 2012 to help address this problem, and to stimulate thinking about researchers' career paths. The Society of Research Software Engineering (\url{https://society-rse.org}) has been established to further promote research software engineer interests.

There is no shortage of computational tools available to help address the problems. However, it should be noted that such tools are not a panacea, as the study \cite{jupyter-study} cited above makes clear. This suggests that human support for improving coding and reproduction quality, perhaps in the form of Working Groups or Boards, as this paper suggests, will be important.

In addition to unintentional problems with code quality and reproducibility, scientific misconduct occurs when the outcome is intentional. While pure plagiarism, which is misconduct, generally does not affect the quality of reported science, when data or code is fraudulently manipulated to have realistic properties, the results are likely to be destructive. The Wakefield MMR fraud linking vaccines and autism published in \emph{The Lancet\/} took 12 years before it was retracted, and has been extraordinarily destructive \cite{MMR-fraud}.

A recent meta-analysis of surveys of scientific misconduct estimates that nearly 2\% of scientists at least once fabricated, falsified, or modified results, and over a third undertook other questionable research practices \cite{misconduct-issues}. The meta-analysis qualifies the figures carefully, but these are alarming rates with any qualifications; indeed, the authors suggest that as misconduct is a sensitive issue, the rates are likely to be under-estimates.\footnote{The author's survey of scientists publishing in the \emph{Journal of Machine Learning\/} had comparable results \cite{machine-learning-reproducibility}.}

While technical solutions like using AI may help, it is notable that many misconduct issues can be detected and constructively managed prior to publication using the same methods as will improve research reproducibility, as discussed throughout the present paper.

\subsection{Computable papers}\label{computable-papers}
``Electronic lab notebooks'' (ELNs) \cite{elns}, emphasize computer tools that specifically support laboratory notebook authoring and editing. In contrast, ``computable papers'' aim to support the scientific paper authoring process: the emphasis is that papers should produce faithful results from code embedded in (or easily accessible from) the text of the paper.

If, for example, HTML is being used to prepare a computational paper, it could include Javascript code like \texttt{<script> document.write(responses.total) </script>}, which would insert the result of running that Javascript code into the paper, such as the number into a sentence in an empirical paper, like ``We collected data and obtained 754 responses.''

Systems like \LaTeX\ (which was used for the current paper) can combine advanced typesetting with basic calculations (e.g., here we calculate $10!=
\def\factorial#1{\ffactorial{#1}1}
\newcount\t
\def\ffactorial#1#2{%
\ifnum#1=0 #2\else
\ffactorial{\the\numexpr#1-1\relax}{\the\numexpr#1*#2\relax}%
\fi}
\factorial{10}$ by writing \LaTeX\ code \emph{in\/} this paper). More practically, a separate program can generate a file of \LaTeX\ definitions for data, tables, and cross-references, etc, as needed for a paper.

\newdimen\boxwidth
\boxwidth = \columnwidth
\advance\boxwidth by - 1em
\noindent\framebox[\columnwidth][r]{\parbox{\boxwidth}{\noindent\sf For the present paper, as a concrete example of this approach, the pilot survey analyzed \textbf{\the\dataN} papers with \textbf{\the\countAuthors} authors. The paper \textbf{\cite{ICmodel}}, discussed below in section \textbf{\ref{section-pandemic-modeling}}, relied on code composed of \textbf{\covidsimfiles} files, with over \textbf{\covidsimkLOC\ thousand} lines of code (how this compares with files in the survey is summarized in table \textbf{\ref{supplementaryTableRepoSummary}} in the \supplement).}}

In the ``old days'' such numbers (\the\dataN, \the\countAuthors, etc) quoted above would have been manually worked out, then eyeballed and typed up by the authors. Instead, in this paper, those numbers (and many more) were computed automatically. Furthermore, they update automatically when the data or calculations change. They were inserted into the text of this paper automatically, and are auditable back to their sources.

The idea is easy to implement. Code can generate text like \texttt{\textbackslash newcommand\{\textbackslash numberOfAuthors\}\{\the\countAuthors\}} (where the \the\countAuthors\ is the calculated number) save that line of text to a file, which is then input into the paper where it is typeset in the normal way. Then, when and wherever the authors write the name \texttt{\textbackslash numberOfAuthors} in their paper's text, the typeset paper says \the\countAuthors, or whatever the correct value is at the time. Some easy \LaTeX\ coding can then present the numbers in the publisher's preferred style, such as \numbertoname 0, \numbertoname 1, \ldots\ \numbertoname 9, \numbertoname{10}, \numbertoname{11} \ldots\ 1,000 etc, as done (for example) in this paper's footnote~\numbernamesexamplefootnote.

Rather than use quirky languages like \LaTeX\ computable papers more often build on general-purpose programming languages. Python and R, for instance, have powerful features for processing data, plotting graphs, doing statistics, and more.

The key concept in computable papers is that as the authors of a paper collect or revise data or calculations or otherwise update it, the results in the paper update automatically, and all the statistics, graphs, and analysis reported in the paper remain correct --- \emph{with no further work from the authors}. Indeed, if an author corrects a mistake, the correction will apply automatically to all future edits.

However, there is no structure to using general-purpose systems like HTML, \LaTeX, or Rmarkdown. Many authors therefore prefer a more structured approach imposed by tools designed for the purpose. A taut review is Perkel \cite{perkel-review}, but here four tools,  WEB, \emph{Mathematica}, Jupyter, and knitr, will serve to sample the variety of approaches that are available:

\begin{itemize}\raggedright

\item
WEB is the earliest tool reviewed here. WEB was developed by Donald Knuth in 1984 \cite{LP,LP-book} as a batch (non-interactive) tool to support his then radical new concept of \emph{literate programming}. The idea was to facilitate programmers write literate documentation for their code. WEB combines a sequential documentation file with code that can be presented in any order, thus overcoming the problem that the best explanation of a program is not necessarily written in the same order as the code it explains. The original WEB allowed Pascal programs to be documented in \TeX, but many variants of WEB have since been developed that are more flexible in the systems they support.

\hskip 1em WEB documents an entire program, but there are variants such as relit \cite{relit} that allow arbitrary parts of programs to be documented, and hence are useful for normal scientific papers that need to explain algorithms, but do not need to show or explain the entire code.

\hskip 1em In contrast to the other tools reviewed here, literate programming is intended to produce high quality publications \emph{about\/} code, rather than publications \emph{using\/} the code, inserting output, such as graphs, generated by running it.

\item
\emph{Mathematica\/} is one of the earliest fully interactive notebook tools for computational papers. Notebooks were developed by Theodore Gray in 1988 \cite{notebook-patent}. Notebooks consist of a collection of ``cells.'' Some notebook cells are labelled as text or as section headings, but if a cell is labelled as code, it can be run and it will normally generate a new cell following it as its output. The new cell can be numbers, tables, mathematics, a plot, an image, or even more text --- the arbitrary output of running code, in fact. Moreover, a notebook itself is a \emph{Mathematica\/} expression, so it --- the paper itself --- can be analyzed or manipulated by code in any way. In particular, the entire notebook structure and contents can be checked for consistency and correctness in any way the author chooses.

\hskip 1em A \emph{Mathematica\/} notebook can be published directly as a paper, but some code might be distracting. Typically the author therefore hides some or all code cells that are irrelevant to the narrative of the paper, but which nonetheless were required to generate the results presented.

\hskip 1em The user manual for \emph{Mathematica\/} itself was written as a \emph{Mathematica\/} notebook, which ensured all its examples actually worked --- and probably helped ensure the correctness of the \emph{Mathematica\/} code used behind the scenes (see section \ref{paper-as-lab}). The book is now the largest example of software documentation in existence: in its latest edition it runs to over 10,000 pages.

\item
Jupyter was developed by Fernando P{\'e}rez and Brian Granger \cite{jupyter-history}, and takes a similar approach to \emph{Mathematica\/}, but is open-source and not closely integrated with any particular programming language, as \emph{Mathematica\/} is. Jupyter can be installed on a PC or run over the web.

\hskip 1em Jupyter is both an authoring tool and a framework on which to build other tools: thus Google's colaboratory is built on top of Jupyter, using it as a foundation but making stylistic changes, including providing free computational resources.

\hskip 1em Many extensive examples of using Jupyter notebooks (and other good practice, such as using repositories) to support large scale science projects can be found at the Gravitational Wave Open Science Center at \url{www.gw-openscience.org}.

\item
knitr \cite{knitr} is a powerful culmination of a variety of tools, Pweave, Sweave, and ideas from literate programming. Knitr combines a markdown document with R code, and is a more powerful approach than the analogous, and perhaps more familiar, HTML+Javascript example shown above to motivate this section.
\end{itemize}

\noindent
Thimbleby \cite{spec-led-design} is a 1999 example of a peer-reviewed paper (about user interface design) written as a \emph{Mathematica\/} notebook, which makes the point that a distinctive feature of its methodology is that the \emph{Mathematica\/} notebook creates a fully inspectable and replicable process. The notebook is available on the author's web site; it can be checked by others, or easily extended or repurposed to support new research --- and it still works 
{\newcount \howlong \howlong = \year
\advance \howlong by -1999
\the\howlong}
years later.

There are tools to make using code more convenient and more reproducible, as this section briefly reviewed, but they are rarely used or used haphazardly, as the next section shows.

\subsection{RAP: Reproducible Analytical Pipelines}\label{RAP-section}
Writing a paper typically starts in a word processor (such as Microsoft Word), sketching an outline, writing boiler-plate text (such as the authors' names and standard section headings), and then gradually building up the evidence base (including citing the literature) that the paper relies on. The process will be concurrent with many other activities --- grant writing, writing up lab books, negotiating authorship, protecting IP, workshops, finding publication outlets, and so on.

Table \ref{fig-pipeline} illustrates the core pipeline of how experiments and data are used to provide information on which analysis and calculations are based, the results of which are then edited into a paper.

\begin{table*}
{\sf
\begin{center}
\def\drop#1{\setbox0=\hbox{\lower .5em\hbox{#1}}%
\ht0=0em
\dp0=0em
\copy0}
\def\lowarrow{\drop{$\rightarrow$}}

\def\bigBracket{$
	\setbox0=\hbox{\lower 1.2ex\hbox{$\left.\vbox to 4em{\vfill}\right\}$}}
	\ht0=0em \dp0=0em 
	~\copy0
$}

\begin{tabular}{|lcl@{}cl@{\hskip 2em}cl|} \hline
\drop{\bfseries Data sources}&\lowarrow&\bfseries \drop{Analysis} & \lowarrow&\bfseries Select results& \lowarrow& \bfseries Submit for\\
\bfseries &&\bfseries & &\bfseries for write up & & \bfseries publication \\ \hline \hline 
&&&&&&\\
Experiments&& Hand calculations &&&&\\
Standard data&& Packages  && Copy \& paste &&\\
Search engines&& SPSS etc & \bigBracket & and edit data && Final paper\\
Literature &&  Graphics packages  && (text, images, graphs, etc)&&\\
Sensors && Specially-written code && into paper &&\\ 
\vdots && \vdots &&&& \\ &&&&&&\\ \hline
\end{tabular}\end{center}}\caption{A simplified schematic of the publication pipeline. For clarity, the pipeline has been linearized; in general, there will be repetitive cyclic iteration and refinement. The RAP and \RAPstar\ approaches encode the normally manual steps in the pipeline processes so that they can be run automatically, and hence reproduce the results that underpin the final paper. The encoded \RAPstar\ algorithms can be shared with other scientists, scrutinized, simplified and optimized, and themselves turned into publishable objects --- they are scientific instruments, just like thermometers or DNA sequencers. Additional schematics are provided in section \ref{RAP-diagrams} in the \supplement.}
\label{fig-pipeline}
\end{table*}

\begin{table*}
{\sf
\newcount \enum
\enum=1
\renewcommand{\theenumi}{%\ifnum \enum<10 \hphantom{0}\fi
\the\enum
\global\advance \enum by 1}
\begin{enumerate}
\item
Peer-review used to ensure the process is reproducible and to identify improvements 
\item
No or minimal manual interference;
for example copy-paste, point-click and drag-drop steps replaced using computer code that can be inspected by others 
\item
Open-source programming languages so that processes do not rely on proprietary software and can be reproduced by others
\item
Version control software, such as Git, to guarantee an audit trail of changes made to code 
\item
Publication of code, whenever possible, on code hosting platforms such as GitHub to improve transparency 
\item
Well-commented code and embedded documentation to ensure the process can be understood and used by others 
\item
Embedding of existing quality assurance practices in code, following guidance set by recognized organizations 
\end{enumerate}
\emph{Adopting RAP principles is not necessarily about incorporating all of the above: implementing just some of these principles will generate valuable improvements.}
}\caption{A minimum standard of RAP, based on the UK Statistics Authority summary \cite{RAPstandard}.}
\label{figure-RAP-summary}
\end{table*}

For clarity, the schematic pipeline in table \ref{fig-pipeline} omits many steps in the creative scientific process. Furthermore, each step is iterated and modified as the research progresses, and, indeed, as referees require revision. The point is that in typical scientific practice each step in the table is largely or entirely manual, typically selecting and copying output from the previous phase, and then pasting and editing the results into the next. The pipeline of data $\rightarrow\cdots\rightarrow$ paper is then iterated by hand as the various components are refined and improved until the authors are happy with the final paper.

As problems are found in a paper, the data, calculations and code are debugged, refactored, and refined. The process is rarely systematic, and even less likely to be documented --- after all, the atomic steps seem to be innocuous copy and paste actions. The final paper and the ideas it embodies are what matters.
